Off the beaten path: \\ Gaia reveals GD-1 stars perturbed off of the main stream track

### 2018-04-30 paper/gd1.tex
trim trim

To study the global properties of the stream as a function of stream longitude,
we extract stream stars around a low-order polynomial model for the median
stream latitude, $\phi_2$, as a function of stream longitude, $\phi_1$.
We find stream ridge points by computing a running median in $\phi_2$ in steps
of $2^\circ$ with a window size of $4^\circ$ along $\phi_1$, then fit a
second-order polynomial to the ridge points to find $\phi_{2,
\textrm{track}}(\phi_1)$.
We then define the stream as the region within $\left| \phi_2 - \phi_{2,
\textrm{track}}(\phi_1) \right| < 0.75^\circ$.
\figurename~\ref{fig:track-and-model}, panel (a), shows the high-confidence
stream stars (same as in lower panel of \figurename~\ref{fig:selection}), and
the two curved lines (blue) show the adopted upper and lower boundary of the
stream region.

We extract properties of the stream from the region defined above as a function
of $\phi_1$ by computing median stream properties in overlapping windows:
we set the window size to $3^\circ$ in $\phi_1$, and shift the window center by
$1^\circ$ in longitude for each subsequent measurement.
Panels (c)--(e) in \figurename~\ref{fig:track-and-model} show the
background-subtracted stream surface density and median proper motion
components, $\mu_{\phi_1}$ and $\mu_{\phi_2}$, extracted in this way (dark
lines), with the associated empirical scatters (shaded regions).
We estimate the local background surface density as an average of areas north
and south of the bounded stream region.

@@

Clear density variations, also apparent in the 2D positions of the sources,
manifest as sharp features in the surface density estimates along the stream.
These are not a result of variable dust extinction or completeness variations
from the \gaia\ scanning pattern.
First, the visible portions of the GD-1 stream are located at high Galactic
latitudes ($b > 20^\circ$), and we therefore do not expect significant dust
extinction or variations in extinction within the stream region.
For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
1^\circ$, the median $V$-band extinction is $\textrm{med}\left(A_V\right) =
0.04~\textrm{mag}$, and the dispersion in $A_V$ in this region is $\sigma_{A_V}
\approx 0.03~\textrm{mag}$, as computed from the SFD
extinction map (\citealt{Schlegel:1998}).
For longitudes $\phi_1 < -60^\circ$, as the stream approaches the Galactic disk,
dust extinction becomes more appreciable, and the stream stars are harder to
select apart from the background.
Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
periods (median and root-variance of 14 and 3, respectively), guaranteeing a
robust estimate of proper motions.
The variance in the number density of faint stars originating from the \gaia\
scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
\mu_{\phi_2, \odot} < 4~\masyr$ (i.e. above the stream selection box), far
smaller than the density variations in the stream.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they

The density variations likely represent real morphological changes along the
GD-1 stream, as have been suggested before from binned samples of
color-magnitude-selected GD-1 stars (\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream.
There is also a well-defined, sharp overdensity in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\todo{cite who}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material that has not yet phase-mixed in multiple dimensions.

We also measured the median proper motion profile along the stream (panels
(d)--(e)) and find that these are consistent with previously measured mean
proper motions from SDSS-USNOB (\citealt{Koposov:2010}).
Here, the proper motion gradient is resolved in striking detail for individual
stars from \gaia\ (see, e.g., elongated overdensity in upper right panel of
\figurename~\ref{fig:selection}).

We compare the measured stream properties to a simple model for the phase-space
density of the stream, generated by only simulating the orbital evolution of
cluster stars once they are tidally stripped from the progenitor.
We use the ``particle-spray'' stream generation method (\citealt{Fardal:2015})
and assume a uniform mass-loss history to generate the model stream, which
choices matches results from more detailed ($N$-body) numerical experiments of
globular cluster disruption \citep[e.g.,][]{Kupper:2012}.
% During simulation, star particles are released from the Lagrange points of the
% cluster with a spread in position and velocity set by the mass of the
% satellite and the location within the host galaxy, with some tunable scale
% parameters (using the parametrization and adopted scale parameters of
% \cite{Fardal:2015}).  This method can accurately reproduce the mean track of
% the resulting stream, but the density of stars along the stream depends in
% detail on the mass-loss history and internal kinematics of the progenitor
% system.  Here we assume a constant mass-loss rate, and release stream
% particles uniformly in time from each Lagrange point; these choices match
% results from more detailed numerical experiments of globular cluster
% disruption \citep[e.g.,][]{Kupper:2012}.

To compute initial conditions for the stream model, we fit an orbit to the
observed properties of the stream in a fixed model for the gravitational
potential of the Milky Way.
We use a three-component potential model to represent the Galaxy, consisting of
a disk (\citealt{Miyamoto:1975}), a bulge
(\citealt{Hernquist:1990}), and a spherical dark matter halo
(\citealt{Navarro:1996}).
We set the disk scale-length parameters to match the disk model found in
\citet{Bovy:2015}; other parameters are set to produce an approximately flat
disk rotation curve between with a circular velocity at the solar circle,
$v_{\textrm{circ}, \odot} \approx 230~\kms$.
We use L-BFGS-B optimization as implemented in the \package{scipy}
\texttt{Python} package to maximize the likelihood of the orbit given sky track
and proper motions from this work with mean distance and giant star radial
velocities from \cite{Koposov:2010}.
We do not vary the parameters of the Milky Way model in the fit because we are
only interested in producing a comparison model of the stream properties.

>>>>

To study the global properties of the stream as a function of stream longitude,
we extract stream stars around a second-order polynomial model for the stream
latitude, $\phi_2$, as a function of stream longitude, $\phi_1$.
We find the stream midpoints by computing the median $\phi_2$ in $4^\circ$
windows shifted by $2^\circ$ along $\phi_1$, then fit a parabola to the median
positions to find $\phi_{2, \textrm{track}}(\phi_1)$.
We then define the stream region as $\left| \phi_2 - \phi_{2,
\textrm{track}}(\phi_1) \right| < 0.75^\circ$.
\figurename~\ref{fig:track-and-model}, panel (a), shows the high-confidence
stream stars (same as \figurename~\ref{fig:selection}), and the two curved lines
(blue) show the upper and lower boundary of the stream region.

We extract properties of the stream from the region defined above as a function
of $\phi_1$ by computing median stream properties in overlapping windows:
we set the window size to $3^\circ$ in $\phi_1$, and shift the window center by
$1^\circ$ in longitude for each subsequent measurement.
Panels (c)--(e) in \figurename~\ref{fig:track-and-model} show the
background-subtracted stream surface density and median proper motion components
extracted in this way (dark lines), with the associated empirical scatters
(shaded regions).
We estimate the local background surface density as an average of areas north
and south of the bounded stream region.
The median proper motion profile along the stream (panels (d)--(e)) are
consistent with previously measured mean proper motions from SDSS-USNOB
(\citealt{Koposov:2010}), but here the proper motion gradient is resolved for
individual stars (see, e.g., elongated overdensity in upper right panel of
\figurename~\ref{fig:selection}).

@@

Clear density variations, also apparent in the 2D positions of the sources,
manifest as sharp features in the surface density estimates along the stream.
These are not a result of variable dust extinction or completeness variations
from the \gaia\ scanning pattern.
First, the visible portions of the GD-1 stream are located at high Galactic
latitudes ($b > 20^\circ$), and we therefore do not expect significant dust
extinction or variations in extinction within the stream region.
The median $V$-band extinction over the stream track is
$\textrm{med}\left(A_V\right) = 0.04~\textrm{mag}$ with a dispersion
$\sigma_{A_V} \approx 0.03~\textrm{mag}$, as computed from the SFD extinction
map (\citealt{Schlegel:1998}).
% As the stream approaches the Galactic disk ($\phi_1 < -60^\circ$),
% dust extinction becomes appreciable, and the stream stars are harder to
% select apart from the background.
Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
periods (median and root-variance of 14 and 3, respectively), promising a robust
estimate of the proper motions.
The variance in the number density of faint stars originating from the \gaia\
scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
\mu_{\phi_2, \odot} < 4~\masyr$, far smaller than the density variations in the
stream.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they

The observed density variations are therefore likely real morphological changes
along the GD-1 stream, as have been suggested before photometric selection alone
(\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream.
There is also a well-defined, sharp overdensity in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\todo{cite who}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material that has not yet phase-mixed in multiple dimensions.

We compare the measured stream properties to a simple model for the phase-space
density of the stream, generated by simulating the orbital evolution of cluster
stars once they are tidally stripped from the progenitor.
We use the ``particle-spray'' stream generation method (\citealt{Fardal:2015})
and assume a uniform mass-loss history to generate the model stream, which
matches results from more detailed ($N$-body) numerical experiments of globular
cluster disruption \citep[e.g.,][]{Kupper:2012}.
% During simulation, star particles are released from the Lagrange points of the
% cluster with a spread in position and velocity set by the mass of the
% satellite and the location within the host galaxy, with some tunable scale
% parameters (using the parametrization and adopted scale parameters of
% \cite{Fardal:2015}).  This method can accurately reproduce the mean track of
% the resulting stream, but the density of stars along the stream depends in
% detail on the mass-loss history and internal kinematics of the progenitor
% system.  Here we assume a constant mass-loss rate, and release stream
% particles uniformly in time from each Lagrange point; these choices match
% results from more detailed numerical experiments of globular cluster
% disruption \citep[e.g.,][]{Kupper:2012}.

To compute initial conditions for the stream model, we fit an orbit to the
observed properties of the stream in a fixed, three-component potential model
to represent the Milky Way, consisting of a disk (\citealt{Miyamoto:1975}), a
bulge (\citealt{Hernquist:1990}), and a spherical dark matter halo
(\citealt{Navarro:1996}).
We set the disk scale-length parameters to match the disk model found in
\citet{Bovy:2015}; other parameters are set to produce an approximately flat
disk rotation curve between with a circular velocity at the solar circle,
$v_{\textrm{circ}, \odot} \approx 230~\kms$.
We use L-BFGS-B optimization as implemented in the \package{scipy}
\texttt{Python} package to maximize the likelihood of the orbit given sky track
and proper motions from this work with mean distance and giant star radial
velocities from \cite{Koposov:2010}.
We do not vary the parameters of the Milky Way model in the fit because we are
only interested in producing a comparison model of the stream properties.
